A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The derivation rules operate on a set $\Ec$ of equalities and disequalities
between $\Sig$-terms, where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
are simplified to $t_k$. All constraints added to $\Ec$ are also implicitly
normalized in this manner.

@@

@@

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{y}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu {\Varec{x}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}}, {\Varec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ {\Varec{x}} \mapsto {\Varec{\const{C}(y,x)}}, {\Varec{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of ${\Varec{y}}$.
\qed
\end{example}

@@

\begin{lemma}[Refutation Soundness]%
\label{lem:rs}%
\afterDot
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

@@

We now define a class of models that we consider.
\rem{Clarify that non-codatatype terms are interpreted as $\mu$-terms, although dropping the $\mu$'s gives the standard representation?}

\begin{definition}[Normal Model] \label{def:norm-model}
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item 
for all types $\tau$, 
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item 
for all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value 
in $\M( \tau )$ that is a renaming of 
$\nf{(\mu x. \const{C}( \M( \vec t \,) ))}$ where $x$ is fresh, and
\item 
for all selector terms $s^j_i( t )$ of type $\tau$,% for some $\vec u$,
$\M( s^j_i( t ) )$ is the value
in $\M( \tau )$ that is a renaming of 
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\mu x. \const{C}_i( \vec u )$.
\end{enumerate}
\end{definition}

@@

First, we need the following definition.
We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.
For a variable $x$ of type $\tau$ and a fixed normal model $\M$,
we write $\mathcal{T}^\M_x( \Val )$ to denote the set consisting of all values $v$ from $\M( \tau )$
such that $v \vsim_x t_1\interp{t_2}$ for some term $t_1$ with subterm $t_2$ occurring in the range of $\Val$.
We first construct a mapping $\Val'$ by exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in \FV( \Val ) 
  \quad
  (\mu x. t) \sim v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_x(\Val)
}{
  \Val := \Val \{ x \mapsto \mu x. t \}
}
\)

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not bisimilar to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $\FV( \Val )$ and does not add any additional variables to $\FV( \Val )$, 
it can only be applied a finite number of times.
We then let $\Val^\ast \ec{x}$ be $\nf{(\Val' \ec{x})}$ for all $\ec{x}$.

@@

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \mu y. \const{B}( x )),
\mu x. \const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \mu y. \const{A}( x )),
\mu x. \const{A}( \mu y. \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu {\Varec{y_2}}. \const{B}( \mu {\Varec{x}}. \const{B}( {\Varec{x}} ) )$, 
which has normal form $\mu {\Varec{y_2}}. \const{B}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu {\Varec{x}}. \const{C}( {\Varec{x}} )$,
$\mu {\Varec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, and
$\mu {\Varec{x}}. \const{B}( \mu x_1. \const{A}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Notice that legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the third substitution mentioned above,
$\Val \ec{w_1}$ is $\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( \mu x_1. \const{A}( {\Varec{x}} ) ) )$.
$\Box$
\end{example}

We first state the following properties of $\Val^\ast$,.
%where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{[#1]}
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\Var}{\mathcal{V}}
%\newcommand{\Varec}[1]{\Var{\ec{#1}}}
\newcommand{\Varec}[1]{\widetilde{#1}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\aequiv}{\mathrel{=_\alpha}}
\newcommand{\vsim}{\aequiv}
\newcommand{\vsimv}[1]{\mathrel{=^{#1}_\alpha}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}
\newcommand{\muvar}{\mathcal{Var}}

@@

The derivation rules operate on a set $\Ec$ of equalities and disequalities
between $\Sig$-terms, where $\Ec$ is interpreted as a conjunction.
We write $\tEc$ to denote the set of $\Sig$-terms occurring in $\Ec$.
%We will commonly denote tuples of terms $( t_1, \ldots, t_n )$ in bold font, as $\vec t$.
%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
%are simplified to $t_k$. All constraints added to $\Ec$ are also implicitly
%normalized in this manner.

@@

%We assume that all $\Sig$-terms in $\Ec$ are \emph{normalized}, meaning that all subterms of the form $\const{s}^{\,k}_j( \const{C}_j(t_1,\ldots,t_n) )$
%are simplified to $t_i$, and moreover assume that all %additional
%constraints added to $\Ec$ are normalized in this manner.
%We present the calculus in three steps.
%In the first step, we compute the bidirectional closure of $\Ec$;
%in the second step, we make inferences based on cyclicity and bisimilarity;
%and in the third step, %(when necessary),
%we branch on constructor types for various terms in $\tEc$.
%Following the conventions from~\cite{}, 
%the derivation rules of our calculus are given in \emph{guarded assignment form},
%where a rule can be applied to $\Ec$ if it meets all of the specified preconditions for $\Ec$.
%The conclusion of a rule either describes equalities to be added to $\Ec$ (in which case we call an application of it \emph{nonterminal}),
%or is $\bot$ (in which case we call an application of it \emph{terminal}).
%A rule may have multiple conclusions separated by $\parallel$, which denotes nondeterministic branching.

@@

\begin{example}
Suppose that $\Ec$ contains distinct equivalence classes $\ec{w}$, $\ec{x}$, $\ec{y}$, and $\ec{z}$
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of \rn{Update} rule applications is given below, omitting
trivial entries such as $\ec{w} \mapsto {\Varec{w}}$.
%
\begin{enumerate}
\item \noindent\rlap{Initial state:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{  \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) \}$
\item \noindent\rlap{Update w.r.t.\ $\ec{y}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) \vthinspace\}$
\item \noindent\rlap{Update w.r.t.\ $\ec{x}$:}\phantom{Update w.r.t.\ $\ec{x}$:\enskip}$\{\vthinspace  \ec{x} \mapsto \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\; \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\, {\Varec{x}} ) ),\;
  \ec{y} \mapsto \mu {\Varec{y}}.\; \const{C}( {\Varec{z}},\; \mu {\Varec{x}}.\; \const{C}( {\Varec{w}},\, {\Varec{y}} ) ) \vthinspace\}$
\end{enumerate}
%
This indicates that the values for $x$ and $y$ in models of $\Ec$
are of the forms $\const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}},\allowbreak \const{C}( {\Varec{z}},\ldots ))))$
and $\const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \const{C}( {\Varec{z}}, \const{C}( {\Varec{w}}, \ldots ))))$,
respectively. %, for some values of ${\Varec{w}}$ and ${\Varec{z}}$.
$\Box$
\end{example}

@@

\begin{example}
Let $\const{C}$ be a constructor of type $\tau_1 \times \tau_2 \rightarrow \tau_2$,
and let $\Ec$ be the set $\{ x \teq \const{C}(y,\const{C}(y,x)) \}$.
After phase~1, the equivalence classes of $\Ec$ are
$\{ x, \const{C}(y,\const{C}(y,x)) \}$, $\{ \const{C}(y,x) \}$, and $\{ y \}$.
We then construct $\Val$ using the aforementioned phases, after which we obtain:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l@{\hspace{.5em}}r@{\hspace{.5em}}l}
\Val \ec{x} & = & 
\mu {\Varec{x}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, {\Varec{x}} ) ) \\
\Val \ec{\const{C}(y,x)} & = & 
\mu {\Varec{\const{C}(y,x)}}. & \const{C}( {\Varec{y}}, & \mu {\Varec{x}}. & \const{C}( {\Varec{y}}, {\Varec{\const{C}(y,x)}} ) )
\end{array}\]
Notice that $\Val \ec{x}$ and $\Val \ec{\const{C}(y,x)}$ are syntactically equivalent for the renaming
$\{ {\Varec{x}} \mapsto {\Varec{\const{C}(y,x)}}, {\Varec{\const{C}(y,x)}} \mapsto {\Varec{x}} \}$,
and thus our calculus may infer $x \teq \const{C}(y,x)$.
Intuitively, this equality holds, since the values of $x$ and $\const{C}( y, x )$ 
are equivalent for any value of ${\Varec{y}}$.
$\Box$
\end{example}

@@

\begin{lemma}[Refutation Soundness]%
\label{lem:rs}%
\afterDot
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is $\thD$-unsatisfiable.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Clash}, or \rn{Acyclic},
then $\Ec$ is $\thD$-unsatisfiable.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Clash}, this is a consequence of distinctness.
For \rn{Acyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is $\thD$-unsatisfiable.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, \rn{Inject}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Sym}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Inject}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is $\thD$-unsatisfiable, 
and thus $\Ec$ is $\thD$-unsatisfiable.
\qed
\end{proof}

@@

We now define a class of models that we consider for a saturated set of equalities $\Ec$.
In the following, we write $\M( \tau )$ to denote the interpretation type $\tau$ in $\M$,
that is, a set of domain elements for that type.
We write $\M( t )$ for a term $t$ to denote the interpretation of $t$ in $\M$,
that is, an element of $\M( \tau )$ where $\tau$ is the type of $t$.
\rem{Clarify that non-codatatype terms are interpreted as $\mu$-terms, although dropping the $\mu$'s gives the standard representation?}

\begin{definition}[Normal Model] \label{def:norm-model}
A model $\M$ is \emph{normal} if and only if:
\begin{enumerate}
\item 
for all types $\tau$, 
$\M( \tau )$ is a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$,
\item 
for all constructor terms $\const{C}( \vec t \,)$ of type $\tau$,
$\M( \const{C}( \vec t \,) )$ is the value 
in $\M( \tau )$ that is $\alpha$-equivalent to
$\nf{(\mu x. \const{C}( \M( \vec t \,) ))}$ where $x$ is fresh, and
\item 
for all selector terms $s^j_i( t )$ of type $\tau$,% for some $\vec u$,
$\M( s^j_i( t ) )$ is the value
in $\M( \tau )$ that is $\alpha$-equivalent to
$\M( t )\interp{u_j}$
when $\M( t )$ is of the form $\mu x. \const{C}_i( \vec u )$.
\end{enumerate}
\end{definition}

@@

First, we need the following definition.
We say that $t \vsimv{x} s$ if and only if $\mu$-terms $t$ and $s$ are syntactically equivalent
for some renaming that avoids capturing any variable other than $x$.
For instance, 
$\mu x. \const{D}( x ) \vsimv{y} \mu x. \const{D}( y )$,
$\mu x. \const{C}( x, x ) \vsimv{x} \mu y. \const{C}( x, y )$ and
$\mu x. \const{C}( z, x ) \vsimv{z} \mu y. \const{C}( z, y )$,
but
$\mu x. \const{D}( x ) \not\vsimv{x} \mu x. \const{D}( y )$ and
$\mu x. \const{C}( x, x ) \not\vsimv{y} \mu y. \const{C}( x, y )$.
For a variable $x$ of type $\tau$ and a fixed normal model $\M$,
we write $\mathcal{T}^\M_x( \Val )$ to denote the set consisting of all values $v$ from $\M( \tau )$
such that $v \vsimv{x} t_1\interp{t_2}$ for some term $t_1$ with subterm $t_2$ occurring in the range of $\Val$.
We construct $\Val^\ast$ by exhaustively applying the following update to $\Val$:

\(
\inferrule{
  x : \tau \in \FV( \Val ) 
  \quad
  (\mu x. t) \aequiv v
  \quad
  v \in \M( \tau )
  \quad
  v \not\in \mathcal{T}^\M_x(\Val)
}{
  \Val := \nf{(\Val \{ x \mapsto \mu x. t \})}
}
\)

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not $\alpha$-equivalent to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $\FV( \Val )$, 
it can only be applied a finite number of times.
After assigning this value, we normalize all terms in the range of $\Val$.
%We then let $\Val^\ast \ec{x}$ be $\nf{(\Val' \ec{x})}$ for all $\ec{x}$.

@@

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to:
\[%\begin{array}{c}
\{
\mu x. \const{A}( \const{B}( x )),
\const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \const{A}( x )),
\mu x. \const{A}( \const{C}( x )),
\mu x. \const{A}( x )
\}
%\end{array}
\]
Now, consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu {\Varec{y_2}}. \const{B}( \mu {\Varec{x}}. \const{B}( {\Varec{x}} ) )$, 
which has normal form $\mu {\Varec{y_2}}. \const{B}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\mu {\Varec{x}}. \const{B}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
then the values in the range of $\Val$ remain $\alpha$-disequivalent.
For instance, we may assign values such as 
$\mu {\Varec{x}}. \const{C}( {\Varec{x}} )$,
$\mu {\Varec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, or
$\mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Notice that legal substitutions for ${\Varec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( \const{A}( {\Varec{x}} ) ) )$,
which has normal form
$\mu {\Varec{w_1}}. \const{A}( \mu {\Varec{x}}. \const{B}( {\Varec{w_1}} ) )$.
$\Box$
\end{example}

We first state the following properties of $\Val^\ast$.
%where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu {\Varec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = {\Varec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.
